---
title: "linear_classification_in_R"
author: "James Le"
date: "9/5/2018"
output: html_document
---

In this post you will discover recipes for 2 linear classification algorithms in R.

All recipes in this post use the iris flowers dataset provided with R in the datasets package. The dataset describes the measurements if iris flowers and requires classification of each observation to one of three flower species.

## Logistic Regression
Logistic Regression is a classification method that models the probability of an observation belonging to one of two classes. As such, normally logistic regression is demonstrated with binary classification problem (2 classes). Logistic Regression can also be used on problems with more than two classes (multinomial), as in this case.

This recipe demonstrates multinomial logistic regression method on the iris dataset.

```{r}
# load the package
library(VGAM)
# load data
data(iris)
# fit model
fit <- vglm(Species~., family=multinomial, data=iris)
# summarize the fit
summary(fit)
# make predictions
probabilities <- predict(fit, iris[,1:4], type="response")
predictions <- apply(probabilities, 1, which.max)
predictions[which(predictions=="1")] <- levels(iris$Species)[1]
predictions[which(predictions=="2")] <- levels(iris$Species)[2]
predictions[which(predictions=="3")] <- levels(iris$Species)[3]
# summarize accuracy
table(predictions, iris$Species)
```

## Linear Discriminant Analysis
LDA is a classification method that finds a linear combination of data attributes that best separate the data into classes.

This recipes demonstrates the LDA method on the iris dataset.

```{r}
# load the package
library(MASS)
data(iris)
# fit model
fit <- lda(Species~., data=iris)
# summarize the fit
summary(fit)
# make predictions
predictions <- predict(fit, iris[,1:4])$class
# summarize accuracy
table(predictions, iris$Species)
```

## Summary
In this post, you discovered 2 recipes for linear classification that you can copy and paste into your own problem.

