---
title: "Linear Regression in R"
author: "James Le"
date: "9/21/2018"
output: html_document
---

In this post you will discover 4 recipes for linear regression for the R platform.

You can copy and paste the recipes in this post to make a jump-start on your own problem or to learn and practice with linear regression in R.

Each example in this post uses the longley dataset provided in the datasets package that comes with R. The longley dataset describes 7 economic variables observed from 1947 to 1962 used to predict the number of people employed yearly.

## Ordinary Least Squares Regression
Ordinary Least Squares (OLS) regression is a linear model that seeks to find a set of coefficients for a line/hyper-plane that minimise the sum of the squared errors.

```{r}
# load data
data(longley)
# fit model
fit <- lm(Employed~., longley)
# summarize the fit
summary(fit)
# make predictions
predictions <- predict(fit, longley)
# summarize accuracy
mse <- mean((longley$Employed - predictions)^2)
print(mse)
```

## Stepwize Linear Regression
Stepwise Linear Regression is a method that makes use of linear regression to discover which subset of attributes in the dataset result in the best performing model. It is step-wise because each iteration of the method makes a change to the set of attributes and creates a model to evaluate the performance of the set.

```{r}
# load data
data(longley)
# fit model
base <- lm(Employed~., longley)
# summarize the fit
summary(base)
# perform step-wise feature selection
fit <- step(base)
# summarize the selected model
summary(fit)
# make predictions
predictions <- predict(fit, longley)
# summarize accuracy
mse <- mean((longley$Employed - predictions)^2)
print(mse)
```

## Principal Component Regression
Principal Component Regression (PCR) creates a linear regression model using the outputs of a Principal Component Analysis (PCA) to estimate the coefficients of the model. PCR is useful when the data has highly correlated predictors.

```{r}
# load the package
library(pls)
# load data
data(longley)
# fit model
fit <- pcr(Employed~., data=longley, validation="CV")
# summarize the fit
summary(fit)
# make predictions
predictions <- predict(fit, longley, ncomp=6)
# summarize accuracy
mse <- mean((longley$Employed - predictions)^2)
print(mse)
```

## Partial Least Squares Regression
Partial Least Squares (PLS) Regression creates a linear model of the data in a transformed projection of problem space. Like PCR, PLS is appropriate for data with highly-correlated predictors.

```{r}
# load the package
library(pls)
# load data
data(longley)
# fit model
fit <- plsr(Employed~., data=longley, validation="CV")
# summarize the fit
summary(fit)
# make predictions
predictions <- predict(fit, longley, ncomp=6)
# summarize accuracy
mse <- mean((longley$Employed - predictions)^2)
print(mse)
```

## Summary
In this post you discovered 4 recipes for creating linear regression models in R and making predictions using those models.