---
title: "Non-Linear Classification in R with Decision Trees"
author: "James Le"
date: "8/29/2018"
output: html_document
---

In this post you will discover 4 recipes for non-linear classification with decision trees in R.

All recipes in this post use the iris flowers dataset provided with R in the datasets package. The dataset describes the measurements if iris flowers and requires classification of each observation to one of three flower species.

## Classification and Regression Trees
Classification and Regression Trees (CART) split attributes based on values that minimize a loss function, such as sum of squared errors.

The following recipe demonstrates the recursive partitioning decision tree method on the iris dataset.

```{r}
# load the package
library(rpart)
# load data
data(iris)
# fit model
fit <- rpart(Species~., data=iris)
# summarize the fit
summary(fit)
# make predictions
predictions <- predict(fit, iris[,1:4], type="class")
# summarize accuracy
table(predictions, iris$Species)
```

## Bagging CART
Bootstrapped Aggregation (Bagging) is an ensemble method that creates multiple models of the same type from different sub-samples of the same dataset. The predictions from each separate model are combined together to provide a superior result. This approach has shown participially effective for high-variance methods such as decision trees.

The following recipe demonstrates bagging applied to the recursive partitioning decision tree for the iris dataset.

```{r}
# load the package
library(ipred)
# load data
data(iris)
# fit model
fit <- bagging(Species~., data=iris)
# summarize the fit
summary(fit)
# make predictions
predictions <- predict(fit, iris[,1:4], type="class")
# summarize accuracy
table(predictions, iris$Species)
```

## Random Forest
Random Forest is variation on Bagging of decision trees by reducing the attributes available to making a tree at each decision point to a random sub-sample. This further increases the variance of the trees and more trees are required.

The following recipe demonstrate the random forest method applied to the iris dataset.

```{r}
# load the package
library(randomForest)
# load data
data(iris)
# fit model
fit <- randomForest(Species~., data=iris)
# summarize the fit
summary(fit)
# make predictions
predictions <- predict(fit, iris[,1:4])
# summarize accuracy
table(predictions, iris$Species)
```

## Boosted C5.0
The C5.0 method is a further extension of C4.5 and pinnacle of that line of methods. It was proprietary for a long time, although the code was released recently and is available in the C50 package.

The following recipe demonstrates the C5.0 with boosting method applied to the iris dataset.

```{r}
# load the package
library(C50)
# load data
data(iris)
# fit model
fit <- C5.0(Species~., data=iris, trials=10)
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, iris)
# summarize accuracy
table(predictions, iris$Species)
```

## Summary
In this post you discovered 4 recipes for non-linear classification using decision trees in R using the iris flowers dataset.

Each recipe is generic and ready for you to copy and paste and modify for your own problem.