---
title: "Spot Check Machine Learning Algorithms in R"
author: "James Le"
date: "1/18/2018"
output: html_document
---

Spot checking machine learning algorithms is how you find the best algorithm for your dataset.

But what algorithms should you spot check?

In this post you discover the 8 machine learning algorithms you should spot check on your data.

You also get recipes of each algorithm that you can copy and paste into your current or next machine learning project in R.

Let’s get started.

There are hundreds of machine learning algorithms available in R.

I would recommend exploring many of them, especially, if making accurate predictions on your dataset is important and you have the time.

Often you don’t have the time, so you need to know the few algorithms that you absolutely must test on your problem.

In this section you will discover the linear and nonlinear algorithms you should spot check on your problem in R. This excludes ensemble algorithms such as as boosting and bagging, that can come later once you have a baseline.

Each algorithm will be presented from two perspectives:

1. The package and function used to train and make predictions for the algorithm.

2. The caret wrapper for the algorithm.

You need to know which package and function to use for a given algorithm. This is needed when:

* You are researching the algorithm parameters and how to get the most from the algorithm.

* You have a discovered the best algorithm to use and need to prepare a final model.

You need to know how to use each algorithm with caret, so that you can efficiently evaluate the accuracy of the algorithm on unseen data using the preprocessing, algorithm evaluation and tuning capabilities of caret.

Two standard datasets are used to demonstrate the algorithms:

* Boston Housing dataset for regression (BostonHousing from the mlbench library).

* Pima Indians Diabetes dataset for classification (PimaIndiansDiabetes from the mlbench library).

Algorithms are presented in two groups:

* Linear Algorithms that are simpler methods that have a strong bias but are fast to train.

* Nonlinear Algorithms that are more complex methods that have a large variance but are often more accurate.

Each recipe presented in this section is complete and will produce a result, so that you can copy and paste it into your current or next machine learning project.

Let’s get started.

## Linear Algorithms
These are methods that make large assumptions about the form of the function being modeled. As such they are have a high bias but are often fast to train.

The final models are also often easy (or easier) to interpret, making them desirable as final models. If the results are suitably accurate, you may not need to move onto non-linear methods if a linear algorithm.

### 1. Linear Regression
The lm() function is in the stats library and creates a linear regression model using ordinary least squares.

```{r}
# load the library
library(mlbench)
# load data
data(BostonHousing)
# fit model
fit <- lm(medv~., BostonHousing)
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, BostonHousing)
# summarize accuracy
mse <- mean((BostonHousing$medv - predictions)^2)
print(mse)
```

The lm implementation can be used in caret as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
# load dataset
data(BostonHousing)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.lm <- train(medv~., data=BostonHousing, method="lm", metric="RMSE", preProc=c("center", "scale"), trControl=control)
# summarize fit
print(fit.lm)
```

### 2. Logistic Regression
The glm function is in the stats library and creates a generalized linear model. It can be configured to perform a logistic regression suitable for binary classification problems.

```{r}
# load the library
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# fit model
fit <- glm(diabetes~., data=PimaIndiansDiabetes, family=binomial(link='logit'))
# summarize the fit
print(fit)
# make predictions
probabilities <- predict(fit, PimaIndiansDiabetes[,1:8], type='response')
predictions <- ifelse(probabilities > 0.5,'pos','neg')
# summarize accuracy
table(predictions, PimaIndiansDiabetes$diabetes)
```

The glm algorithm can be used in caret as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.glm <- train(diabetes~., data=PimaIndiansDiabetes, method="glm", metric="Accuracy", preProc=c("center", "scale"), trControl=control)
# summarize fit
print(fit.glm)
```

### 3. Linear Discriminant Analysis
The lda function is in the MASS library and creates a linear model of a classification problem.

```{r}
# load the libraries
library(MASS)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# fit model
fit <- lda(diabetes~., data=PimaIndiansDiabetes)
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, PimaIndiansDiabetes[,1:8])$class
# summarize accuracy
table(predictions, PimaIndiansDiabetes$diabetes)
```

The lda algorithm can be used in caret as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.lda <- train(diabetes~., data=PimaIndiansDiabetes, method="lda", metric="Accuracy", preProc=c("center", "scale"), trControl=control)
# summarize fit
print(fit.lda)
```

### 4. Regularized Regression
The glmnet function is in the glmnet library and can be used for classification or regression.

Classification Example:

```{r}
# load the library
library(glmnet)
library(mlbench)
# load data
data(PimaIndiansDiabetes)
x <- as.matrix(PimaIndiansDiabetes[,1:8])
y <- as.matrix(PimaIndiansDiabetes[,9])
# fit model
fit <- glmnet(x, y, family="binomial", alpha=0.5, lambda=0.001)
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, x, type="class")
# summarize accuracy
table(predictions, PimaIndiansDiabetes$diabetes)
```

Regression Example:

```{r}
# load the libraries
library(glmnet)
library(mlbench)
# load data
data(BostonHousing)
BostonHousing$chas <- as.numeric(as.character(BostonHousing$chas))
x <- as.matrix(BostonHousing[,1:13])
y <- as.matrix(BostonHousing[,14])
# fit model
fit <- glmnet(x, y, family="gaussian", alpha=0.5, lambda=0.001)
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, x, type="link")
# summarize accuracy
mse <- mean((y - predictions)^2)
print(mse)
```

It can also be configured to perform three important types of regularization: lasso, ridge and elastic net by configuring the alpha parameter to 1, 0 or in [0,1] respectively.

The glmnet implementation can be used in caret for classification as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
library(glmnet)
# Load the dataset
data(PimaIndiansDiabetes)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.glmnet <- train(diabetes~., data=PimaIndiansDiabetes, method="glmnet", metric="Accuracy", preProc=c("center", "scale"), trControl=control)
# summarize fit
print(fit.glmnet)
```

The glmnet implementation can be used in caret for regression as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
library(glmnet)
# Load the dataset
data(BostonHousing)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.glmnet <- train(medv~., data=BostonHousing, method="glmnet", metric="RMSE", preProc=c("center", "scale"), trControl=control)
# summarize fit
print(fit.glmnet)
```

## Nonlinear Algorithms
Thees are machine learning algorithms that make fewer assumptions about the function being modeled. As such, they have a higher variance but are often result in higher accuracy. They increased flexibility also can make them slower to train or increase their memory requirements.

### 1. k-Nearest Neighbors
The knn3 function is in the caret library and does not create a model, rather makes predictions from the training set directly. It can be used for classification or regression.

Classification Example:

```{r}
# knn direct classification
 
# load the libraries
library(caret)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# fit model
fit <- knn3(diabetes~., data=PimaIndiansDiabetes, k=3)
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, PimaIndiansDiabetes[,1:8], type="class")
# summarize accuracy
table(predictions, PimaIndiansDiabetes$diabetes)
```

Regression Example:

```{r}
# load the libraries
library(caret)
library(mlbench)
# load data
data(BostonHousing)
BostonHousing$chas <- as.numeric(as.character(BostonHousing$chas))
x <- as.matrix(BostonHousing[,1:13])
y <- as.matrix(BostonHousing[,14])
# fit model
fit <- knnreg(x, y, k=3)
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, x)
# summarize accuracy
mse <- mean((BostonHousing$medv - predictions)^2)
print(mse)
```

The knn implementation can be used within the caret train() function for classification as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.knn <- train(diabetes~., data=PimaIndiansDiabetes, method="knn", metric="Accuracy", preProc=c("center", "scale"), trControl=control)
# summarize fit
print(fit.knn)
```

The knn implementation can be used within the caret train() function for regression as follows:

```{r}
# load libraries
library(caret)
data(BostonHousing)
# Load the dataset
data(BostonHousing)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.knn <- train(medv~., data=BostonHousing, method="knn", metric="RMSE", preProc=c("center", "scale"), trControl=control)
# summarize fit
print(fit.knn)
```

### 2. Naive Bayes
The naiveBayes function is in the e1071 library and models the probabilistic of each attribute to the outcome variable independently. It can be used for classification problems.

```{r}
# load the libraries
library(e1071)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# fit model
fit <- naiveBayes(diabetes~., data=PimaIndiansDiabetes)
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, PimaIndiansDiabetes[,1:8])
# summarize accuracy
table(predictions, PimaIndiansDiabetes$diabetes)
```

A very similar naive bayes implementation (NaiveBayes from the klaR library) can be used with caret as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.nb <- train(diabetes~., data=PimaIndiansDiabetes, method="nb", metric="Accuracy", trControl=control)
# summarize fit
print(fit.nb)
```

### 3. Support Vector Machine
The ksvm function is in the kernlab package and can be used for classification or regression. It is a wrapper for the LIBSVM library and provides a suite of kernel types and configuration options.

These example uses a Radial Basis kernel.

Classification Example:

```{r}
# load the libraries
library(kernlab)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# fit model
fit <- ksvm(diabetes~., data=PimaIndiansDiabetes, kernel="rbfdot")
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, PimaIndiansDiabetes[,1:8], type="response")
# summarize accuracy
table(predictions, PimaIndiansDiabetes$diabetes)
```

Regression Example:

```{r}
# load the libraries
library(kernlab)
library(mlbench)
# load data
data(BostonHousing)
# fit model
fit <- ksvm(medv~., BostonHousing, kernel="rbfdot")
# summarize the fit
print(fit)
# make predictions
predictions <- predict(fit, BostonHousing)
# summarize accuracy
mse <- mean((BostonHousing$medv - predictions)^2)
print(mse)
```

The SVM with Radial Basis kernel implementation can be used with caret for classification as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
# Load the dataset
data(PimaIndiansDiabetes)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.svmRadial <- train(diabetes~., data=PimaIndiansDiabetes, method="svmRadial", metric="Accuracy", trControl=control)
# summarize fit
print(fit.svmRadial)
```

The SVM with Radial Basis kernel implementation can be used with caret for regression as follows:

```{r}
# load libraries
library(caret)
library(mlbench)
# Load the dataset
data(BostonHousing)
# train
set.seed(7)
control <- trainControl(method="cv", number=5)
fit.svmRadial <- train(medv~., data=BostonHousing, method="svmRadial", metric="RMSE", trControl=control)
# summarize fit
print(fit.svmRadial)
```

# Summary
In this post you discovered a diverse set of 8 algorithms that you can use to spot check on your datasets. Specifically:

* Linear Regression

* Logistic Regression

* Linear Discriminant Analysis

* Regularized Regression

* k-Nearest Neighbors

* Naive Bayes

* Support Vector Machine

* Classification and Regression Trees

You learned which packages and functions to use for each algorithm. You also learned how you can use each algorithm with the caret package that provides algorithm evaluation and tuning capabilities.

You can use these algorithms as a template for spot checking on your current or next machine learning project in R.